# Data manipulation with `dplyr` and `tidyr`
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 90)
```

# Motivation

How much of data science or machine learning is data manipulation?

Advantages of `dplyr` and `tidyr`:

* fast
* flexible
* easy to use and easy to read

`dplyr` is a major galaxy in the tidyverse, while `tidyr` is a specialized star.

# Introductory Examples

## dplyr

```{r, message=FALSE}
library(dplyr)

## data source:
## https://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Medicare-Provider-Charge-Data/Inpatient2014.html
df <- readRDS("data/inpatient_charges_2014_clean_cardiac_50plus.RDS")
```
```{r, eval=FALSE}
## can you guess what this does?
df %>%
    group_by(DRG.code) %>%
    summarise(avg_discharges = mean(Total.Discharges)) %>%
    arrange(desc(avg_discharges))
```

Q: Which of these is easier to read?

```{r, eval=FALSE}
paste0("(", gsub(".", "_", tolower(names(df)), fixed=TRUE), ")")
## or:

names(df) %>%
    tolower() %>%
	gsub(".", "_", ., fixed=TRUE) %>%
	paste0("(", ., ")")

```

## tidyr


```{r}
## reshaping data with `tidyr`
## starting with a "wide" data frame
library(tidyr)
(wide <- df %>% select(DRG.code, Provider.Id, starts_with("Ave")))

## make it "long"
(long <- wide %>% gather(data=., key=Type, value=Amount, 3:5) %>%
   arrange(DRG.code,Provider.Id ))

## back to wide
(back_to_wide  <- long %>% spread(., Type, Amount))
```



# Concepts

`dplyr` is an R package for easier data manipulation of **d**ata frames. `tidyr` is a package for reshaping data into a "tidy" format (or long to wide, and wide to long).

In the `dplyr` workflow, you start with a dataframe, then apply the verbs to do
what you want. The result is a new data frame.

`dplyr` includes the pipe `%>%` operator out of the box, as well as the data frame type `tbl_df`.

What is the pipe? It’s absolutely amazing. The pipe `%>%` “passes the object on the left hand side as the first argument (or . argument) of the function on the right-hand side:”

```{r, eval=FALSE}
x %>% f(y) is the same as f(x, y)

y %>% f(x, ., z) is the same as f(x, y, z )
```

`tbl_df` is like a data frame, except that it has a better default print: it prints only 10 records, and provides data type (like `str()`).

The basic verbs (from the official vignette):

  dplyr aims to provide a function for each basic verb of data manipulation:

  filter() (and slice())
  arrange()
  select() (and rename())
  distinct()                      #remove duplicate rows
  mutate() (and transmute())      #make new columns
  summarise()
  sample_n() (and sample_frac())


Let's look at each:

```{r}
## first, a concise summary of the data
df %>% glimpse
```
```{r, eval=FALSE}
## equivalent to:
glimpse(df)
```

```{r}
## filter() rows
df %>%
    filter(DRG.code==233)

## (can assign the result to a new data frame)
new_df <- df %>%
    filter(DRG.code==233)

## select() columns
df %>%
    select(DRG.code, Provider.Id, Total.Discharges)

## deselect with -
df %>%
    select(-DRG.Definition, -DRG.text)

## arrange() result
df %>%
    filter(DRG.code==233) %>%
    select(Provider.Id, Total.Discharges, Provider.State) %>% 
    arrange(desc(Total.Discharges))

## summarise(), perhaps with a group_by()
## note: summarise() collapses data, as opposed to mutate, which creates new columns
## note: you don't need to select the variables before the summarize
df %>%
    summarise(total=sum(Average.Total.Payments))

df %>%
    group_by(DRG.code) %>%
    summarise(providers=n(),
              total=sum(Average.Total.Payments)) %>%
    arrange(desc(total))

## make new columns with mutate()
df %>%
    group_by(Provider.Id) %>%
    select(starts_with("Ave")) %>%
    mutate(Ave_max=max(Average.Medicare.Payments, Average.Total.Payments,
                       Average.Covered.Charges))
                      
```

Other capabilities:

```{r}

## n() and n_distinct()
df %>%
    group_by(DRG.code) %>%
    summarise(states = n_distinct(Provider.State),
              codes = n())              #careful with n() 

## compute statistics with summarise()
df %>%
    filter(DRG.code==247) %>%
    group_by(Provider.State) %>% 
    summarise(min_discharges=min(Total.Discharges),
              median_discharges=median(Total.Discharges),
              max_discharges=max(Total.Discharges)) %>%
    arrange(desc(median_discharges))


## obtain distinct values
df %>%
    distinct(Provider.Id)


## bin data into buckets with ntile
df %>%
    select(DRG.code, Provider.Id, Average.Total.Payments) %>%
    group_by(DRG.code) %>%
    mutate(payment_quartile = ntile(Average.Total.Payments, 4))

## subsample data (useful)
df %>%
  sample_n(10)

df %>%
  sample_frac(0.2)
    
```


## Example: an advanced reshape to produce a summary table:

```{r}
df %>% 
    select(starts_with("Average")) %>% 
    summarise_each(funs(n(), mean, sd, min, max))  %>%
    gather(variable, value)  %>%
    separate(variable, c("var", "stat"), sep="\\_") %>%
    spread(var, value)
```


Last, before we do some exercises, let's review this super-useful handout:

[https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)


# Exercises

1. Explore the dataframe `df` with the basic verbs. For example, select a column and find some summary statistics.

2. Find a few ways to select the columns that refer to location.

2. Build a summary table of Total.Medicare.Payment by Provider-state, arranged in descending order.

3. Subset `df` to 10% of the data, and repeat #2.

4. By DRG.code, return top 3 providers (by Total Medicare Payments).

5. Create a target y: 0/1, where 1 means Average.Total.Payments is greater than or equal to its mean across each DRG.code, and 0 otherwise.


# Summary and Takeaways

Download and save/print this data wrangling handout:

[https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

For other useful "window" functions: see the above Data Wrangling handout.


## References
James et al, p. 47

`> vignette("introduction", "dplyr")`

http://stackoverflow.com/documentation/r/topicse


# Advanced/Details

Learn to use `top_n()` to return top N values, perhaps by group.

Why chaining?
https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html#chaining

The basic pipe `%>%` is just the beginning:

```{r, eval=FALSE}
library(magrittr)
## %$%
table(df$DRG.code, df$DRG.text)
## vs
df %$% table(DRG.code, DRG.text)

## %<>% compound operator
temp_df <- df
names(temp_df) %<>% tolower
```

Learn the rest, and see some neat uses of them all on
the Github page for magrittr at [https://github.com/smbache/magrittr](https://github.com/smbache/magrittr)

Example selecting variables programmatically with standard evaluation vs Non-standard
evaluation (SE vs. NSE)

NSE = basically, for less typing. 

Verbs in `dplyr` can end ing `_`. For example: `select()` vs `select_()` etc.

See [http://adv-r.had.co.nz/Computing-on-the-language.html](http://adv-r.had.co.nz/Computing-on-the-language.html)

```{r}
codes  <- c(210:250)
myvars <- c("Provider.Id", "DRG.code", "Average.Medicare.Payments",
            "Average.Covered.Charges", "Average.Total.Payments",
            "Total.Discharges")

dfx <- df %>% filter(DRG.code %in% codes) %>%
    select_(.dots=myvars)
```



How `dplyr` plugs into databases:
https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html#databases

Reshaping data from wide to long or vice-versa is a common task, but I was not satisfied with the docs for spread and gather. So I still refer to this gather and spread cookbook on cookbook-r.com: [http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/)
