# Visualizing data with ggplot2
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 90)
```

# Motivation
To build good machine learning models, you should understand your data. This is even more important if you will be presenting your model results to a non-technical (business) audience.

A good graphic or visualization can be very convincing. 

`ggplot` is the standard for exploratory graphcis in R, and is very popular for publishing too.

# Introductory Examples

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
df0 <- readRDS("data/inpatient_charges_2014_clean_cardiac_50plus.RDS")
```
```{r}
df <- df0 %>% filter(DRG.code %in% c(220:240))
```
```{r, echo=FALSE}
theme_set(theme_bw())
```
```{r}
ggplot(df, aes(x=reorder(factor(DRG.code), Average.Total.Payments, median, order=TRUE), y=Average.Total.Payments, fill=factor(DRG.code))) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha=0.08) +
  coord_flip() +
  scale_y_continuous(breaks=scales::pretty_breaks(8)) +
  ggtitle("Average Total Payments, by DRG code") +
  labs(x="DRG code", y="Payment ($)") +
  theme(legend.position="none")
```


# Concepts

	"ggplot2 is designed to work in a layered fashion, starting with a
	layer showing the raw data then adding layers of annotation and
	statistical summaries."


Plot with `ggplot()` or quick plot with `qplot()`

Specify x, y and aesthetics with `aes()` inside `ggplot()`. Pass string names with `aes_string()`.

Add layers with `+` such as geom's: `+ geom_point()`
```{r, echo=FALSE}
theme_set(theme_bw())
```

```{r}
## Let's illustrate these concepts with a simple scatterplot
ggplot(data=df, aes(x=Average.Covered.Charges, y=Average.Total.Payments)) +
    geom_point()

## easily add smoother 
ggplot(data=df, aes(x=Average.Covered.Charges, y=Average.Total.Payments)) +
  geom_point() +
  geom_smooth(method="lm")

## customize color
ggplot(data=df, aes(x=Average.Covered.Charges, y=Average.Total.Payments)) +
    geom_point(colour="red")

## customize group/color
## use pretty scales with brewer
ggplot(data=df, aes(x=Average.Covered.Charges, y=Average.Total.Payments,
                    colour=factor(DRG.code))) +
    geom_point(alpha=0.4) +
    scale_color_brewer(type="div")

## different geom: a histogram (univariate summary)
ggplot(df, aes(Average.Total.Payments)) + geom_histogram()

## a bar chart
ggplot(df, aes(x=DRG.code)) + geom_bar()
```

## Let's explore ggplot layering with a boxplot
```{r}
## distinct(df, DRG.code)

p <- ggplot(df, aes(x=factor(DRG.code), y=Average.Total.Payments))

p + geom_boxplot()

p + geom_boxplot() + geom_jitter()

p + geom_boxplot(outlier.shape = NA) + geom_jitter(alpha=0.2)

p + geom_boxplot(outlier.shape = NA) + geom_jitter(alpha=0.2) + coord_flip()
```


## Use *facets* to explore data by sub-groups
```{r}
df2 <- df %>% filter(Provider.State %in% c("DC", "VA", "MD"))

ggplot(df2, aes(x=factor(Provider.State), y=Average.Total.Payments)) +
    geom_jitter() +
    facet_wrap(~DRG.code) + 
    ggtitle("Payments by Provider State") +
    labs(y="Average Total Payment", x="State")

ggplot(df2, aes(x=factor(Provider.State), y=Average.Total.Payments)) +
    geom_jitter() +
    facet_grid(~DRG.code) + 
    ggtitle("Payments by Provider State") +
    labs(y="Average Total Payment", x="State")


```


## qplot(): a quick plot, like base R plot()

```{r}
## plots histogram if given 1 variable
qplot(Average.Total.Payments, data=df)

## plots scatterplot if given 2 variables
qplot(Average.Covered.Charges, Average.Total.Payments, data=df)

qplot(Average.Covered.Charges, Average.Total.Payments, data=df,
      geom="line")
```


# Summary

ggplot is the standard for exploratory data visualizations

## References:
James et al, p 45, 49

[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)

[https://github.com/AmeliaMN/ggplot2_tutorial](https://github.com/AmeliaMN/ggplot2_tutorial)

# Exercises

1. Subset data frame `df2` to DC and plot Average.Medicare.Payments for each DRG.code.

2. Use a search engine to find a way to create a scatterplot matrix of the 3 Payment variables. 
